Detection and replication of epistasis influencing transcription in humans

### 2013-12-19 docs/manuscript/shorter.tex
sent revised manuscript to nature

\documentclass{article}
\usepackage[superscript,biblabel]{cite}
\usepackage{graphicx}
\usepackage{float}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage{authblk}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{lscape}
\usepackage{lineno}
\usepackage{hyperref}
\linenumbers

@@

\author[1,2,*]{Gibran Hemani}
\author[1,2]{Konstantin Shakhbazov}
\author[3]{Harm-Jan Westra}
\author[4,5,6]{Tonu Esko}
\author[7]{Anjali K Henders}
\author[1,2]{Allan F McRae}
\author[2]{Jian Yang}
\author[8]{Greg Gibson}
\author[7]{Nicholas G Martin}
\author[4]{Andres Metspalu}
\author[3]{Lude Franke}
\author[7,+]{Grant W Montgomery}
\author[1,2,+]{Peter M Visscher}
\author[1,2,+]{Joseph E Powell}

\affil[1]{University of Queensland Diamantina Institute, University of Queensland, Princess Alexandra Hospital, Brisbane, Queensland, Australia}
\affil[2]{Queensland Brain Institute, University of Queensland, Brisbane, QLD, Australia}
\affil[3]{Department of Genetics, University Medical Center Groningen, University of Groningen, Hanzeplein 1, Groningen, the Netherlands}
\affil[4]{Estonian Genome Center, University of Tartu, Tartu, 51010, Estonia}
\affil[5]{Medical and Population Genetics, Broad Institute, Cambridge, MA, 02142, US}
\affil[6]{Divisions of Endocrinology, Children's Hospital, Boston, MA, 02115, US}
\affil[7]{Queensland Institute of Medical Research, Brisbane, Queensland, Australia}
\affil[8]{School of Biology and Centre for Integrative Genomics, Georgia Institute of Technology, Atlanta, Georgia United States of America}
\affil[+]{These authors contributed equally}
\affil[*]{Corresponding author: g.hemani@uq.edu.au}

@@

In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions where $p$-values are assumed to be uniformly distributed ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}). Concordance of the direction of the effect of the largest variance component was also highly significant ($p = 5.71 \times 10^{-31}$, Supplementary Table \ref{tab:sign}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}). It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistasis is difficult in practice due to LD (Methods).

@@

% \subsection*{Acknowledgements}

% We are grateful to the volunteers for their generous participation in these studies. We thank Bill Hill, Chris Haley and Lars Ronnegard for helpful discussions and comments.

% This work could not have been completed without access to high performance GPGPU compute clusters. We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au). We also thank Jake Carroll and Irek Porebski from the Queensland Brain Institute Information Technology Group for HPC support.

% The University of Queensland group is supported by the Australian National Health and Medical Research Council (NHMRC) grants 389892, 496667, 613601, 1010374 and 1046880, the Australian Research Council (ARC) grant (DE130100691), and by National Institutes of Health (NIH) grants GM057091 and GM099568.

% The QIMR researchers acknowledge funding from the Australian National Health and Medical Research Council (grants 241944, 389875, 389891, 389892, 389938, 442915, 442981, 496739, 496688 and 552485), the and the National Institutes of Health (grants AA07535, AA10248, AA014041, AA13320, AA13321, AA13326 and DA12854). We thank Anthony Caracella and Lisa Bowdler for technical assistance with the micro-array hybridisations.

% The CHDWB study funding support from the Georgia Institute of Technology Research Foundation. The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript

% The Fehrmann study was supported by grants from the Celiac Disease Consortium (an innovative cluster approved by the Netherlands Genomics Initiative and partly funded by the Dutch Government (grant BSIK03009), the Netherlands Organization for Scientific Research (NWO-VICI grant 918.66.620, NWO-VENI grant 916.10.135 to L.F.), the Dutch Digestive Disease Foundation (MLDS WO11-30), and a Horizon Breakthrough grant from the Netherlands Genomics Initiative (grant 92519031 to L.F.). This project was supported by the Prinses Beatrix Fonds, VSB fonds, H. Kersten and M. Kersten (Kersten Foundation), The Netherlands ALS Foundation, and J.R. van Dijk and the Adessium Foundation. The research leading to these results has received funding from the European Community’s Health Seventh Framework Programme (FP7/2007-2013) under grant agreement 259867.

% The EGCUT study received targeted financing from Estonian Government SF0180142s08, Center of Excellence in Genomics (EXCEGEN) and University of Tartu (SP1GVARENG). We acknowledge EGCUT technical personnel, especially Mr V. Soo and S. Smit. Data analyzes were carried out in part in the High Performance Computing Center of University of Tartu.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=2.5in]{gpbonfrep.pdf}
	\caption{\textbf{Replication of GP maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown. Each GP map consists of nine tiles where each tile represents the expression level for that two-locus genotype class. Phenotypes are for gene transcript levels (dark coloured tiles = high expression, light coloured tiles = low expression). Columns of GP maps are for each independent dataset. Rows of GP maps are for each of 30 significantly replicated interactions at the Bonferroni level, corresponding to the rows in Table \ref{tab:bonferroni}. There is a clear trend of the GP maps replicating across all three datasets.}
	\label{fig:gpmaps}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=2.5in]{qqMeta.pdf}
	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} The top panel shows all 434 discovery SNPs that were tested for interactions. Observed $p$-values ($y$-axis, $-\log_{10}$ scale) are plotted against the expected $p$-values ($x$-axis, $-\log_{10}$ scale). The multiple testing correction threshold for significance following Bonferroni correction is denoted by a dotted line. The bottom panel shows the same data as the top panel but excluding the 30 interactions that were significant at the Bonferroni level in the replication datasets. The shaded grey area represents the 5\% confidence interval for the expected distribution of $p$-values. Dark blue points represent $p$-values that exceed the confidence interval, light blue are within the confidence interval.}
	\label{fig:qqMeta}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes). Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data The remaining 89 interactions that did not replicate are depicted in grey. It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations. An interactive version of this graph can be found here: \url{http://kn3in.github.io/detecting_epi/}}
	\label{fig:fireworks}
\end{figure}

@@

\section*{Supplementary Figures}
\setcounter{figure}{0}
\makeatletter 
\renewcommand{\thefigure}{S\@arabic\c@figure} 
\makeatletter

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{qqMetaNonsig}
	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets, excluding the 30 points significant at the Bonferroni level} The right panel (Null) shows the interaction $p$-values from a meta analysis across two independent datasets on 434 randomly drawn SNP pairs. The left panel (Empirical) shows the interaction $p$-values from the 404 putative interactions that were not significant at the Bonferroni correction threshold. Dark blue points represent $p$-values that surpass the 2.5\% FDR level, as in Figure \ref{fig:qqMeta}.}
\label{fig:qqMetaNonsig}
\end{figure}
\clearpage

>>>>

\documentclass{article}
\usepackage[superscript,biblabel]{cite}
\usepackage{graphicx}
\usepackage{float}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage{authblk}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{lscape}
% \usepackage{lineno}
\usepackage{hyperref}
% \linenumbers

@@

\author[1,2,*]{Gibran Hemani}
\author[1,2]{Konstantin Shakhbazov}
\author[3]{Harm-Jan Westra}
\author[4,5,6]{Tonu Esko}
\author[7]{Anjali K Henders}
\author[1,2]{Allan F McRae}
\author[1]{Jian Yang}
\author[8]{Greg Gibson}
\author[7]{Nicholas G Martin}
\author[4]{Andres Metspalu}
\author[3]{Lude Franke}
\author[7,+]{Grant W Montgomery}
\author[1,2,+]{Peter M Visscher}
\author[1,2,+]{Joseph E Powell}

\affil[1]{Queensland Brain Institute, University of Queensland, Brisbane, QLD, Australia}
\affil[2]{University of Queensland Diamantina Institute, University of Queensland, Princess Alexandra Hospital, Brisbane, Queensland, Australia}
\affil[3]{Department of Genetics, University Medical Center Groningen, University of Groningen, Hanzeplein 1, Groningen, the Netherlands}
\affil[4]{Estonian Genome Center, University of Tartu, Tartu, 51010, Estonia}
\affil[5]{Medical and Population Genetics, Broad Institute, Cambridge, MA, 02142, US}
\affil[6]{Divisions of Endocrinology, Children's Hospital, Boston, MA, 02115, US}
\affil[7]{Queensland Institute of Medical Research, Brisbane, Queensland, Australia}
\affil[8]{School of Biology and Centre for Integrative Genomics, Georgia Institute of Technology, Atlanta, Georgia United States of America}
\affil[+]{These authors contributed equally}
\affil[*]{Corresponding author: g.hemani@uq.edu.au}

@@

In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions where $p$-values are assumed to be uniformly distributed ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}). Concordance of the direction of the effect of the largest variance component was also highly significant ($p = 5.71 \times 10^{-31}$, Supplementary Table \ref{tab:sign}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}). It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistatic effects in independent samples is difficult in practice due to LD (Methods).

@@

\section*{Methods Summary}
We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs. Each individual had gene expression levels measured in peripheral blood at 7,339 probes representing 6,158 RefSeq genes (significant expression in $\geq 90\%$ of individuals). SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms. We used permutation analysis to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER). All SNP pairs with LD $r^2 > 0.1$ and $D'^{2} > 0.1$ were removed to minimise the possibility of haplotype effects. All SNP pairs were required to have at least five data points in all nine genotype classes. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained. Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair. The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood Fehrmann \cite{Fehrmann2011}, $n=1240$; EGCUT \cite{Metspalu2004}, ($n=891$). A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.

\subsection*{Acknowledgements}

We are grateful to the volunteers for their generous participation in these studies. We thank Bill Hill, Chris Haley and Lars Ronnegard for helpful discussions and comments.

This work could not have been completed without access to high performance GPGPU compute clusters. We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au). We also thank Jake Carroll and Irek Porebski from the Queensland Brain Institute Information Technology Group for HPC support.

The University of Queensland group is supported by the Australian National Health and Medical Research Council (NHMRC) grants 389892, 496667, 613601, 1010374 and 1046880, the Australian Research Council (ARC) grant (DE130100691), and by National Institutes of Health (NIH) grants GM057091 and GM099568.

The QIMR researchers acknowledge funding from the Australian National Health and Medical Research Council (grants 241944, 389875, 389891, 389892, 389938, 442915, 442981, 496739, 496688 and 552485), the and the National Institutes of Health (grants AA07535, AA10248, AA014041, AA13320, AA13321, AA13326 and DA12854). We thank Anthony Caracella and Lisa Bowdler for technical assistance with the micro-array hybridisations.

The CHDWB study funding support from the Georgia Institute of Technology Research Foundation. The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript

The Fehrmann study was supported by grants from the Celiac Disease Consortium (an innovative cluster approved by the Netherlands Genomics Initiative and partly funded by the Dutch Government (grant BSIK03009), the Netherlands Organization for Scientific Research (NWO-VICI grant 918.66.620, NWO-VENI grant 916.10.135 to L.F.), the Dutch Digestive Disease Foundation (MLDS WO11-30), and a Horizon Breakthrough grant from the Netherlands Genomics Initiative (grant 92519031 to L.F.). This project was supported by the Prinses Beatrix Fonds, VSB fonds, H. Kersten and M. Kersten (Kersten Foundation), The Netherlands ALS Foundation, and J.R. van Dijk and the Adessium Foundation. The research leading to these results has received funding from the European Community’s Health Seventh Framework Programme (FP7/2007-2013) under grant agreement 259867.

The EGCUT study received targeted financing from Estonian Government SF0180142s08, Center of Excellence in Genomics (EXCEGEN) and University of Tartu (SP1GVARENG). We acknowledge EGCUT technical personnel, especially Mr V. Soo and S. Smit. Data analyzes were carried out in part in the High Performance Computing Center of University of Tartu.

\subsection*{Author contributions}

G.H., J.E.P., P.M.V., and G.W.M. conceived and designed the study. G.H., J.E.P., K.S., H-J.W., and J.Y. performed the analysis. T.E. and A.M. provided the EGCUT data. A.K.H., A.F.M., G.W.M., N.G.M., and J.E.P. provided the BSGS data. H-J.W. and L.F. provided the Fehrmann data. G.H. and J.E.P. wrote the manuscript with the participation of all authors.

\subsection*{Author information}

The authors declare no financial competing interests.

@@

\begin{figure}[H]
	\centering
	% \includegraphics[width=2.5in]{gpbonfrep.pdf}
	\caption{\textbf{Replication of GP maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown. Each GP map consists of nine tiles where each tile represents the expression level for that two-locus genotype class. Phenotypes are for gene transcript levels (dark coloured tiles = high expression, light coloured tiles = low expression). Columns of GP maps are for each independent dataset. Rows of GP maps are for each of 30 significantly replicated interactions at the Bonferroni level, corresponding to the rows in Table \ref{tab:bonferroni}. There is a clear trend of the GP maps replicating across all three datasets.}
	\label{fig:gpmaps}
\end{figure}

\begin{figure}[H]
	\centering
	% \includegraphics[width=2.5in]{qqMeta.pdf}
	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} The top panel shows all 434 discovery SNPs that were tested for interactions. Observed $p$-values ($y$-axis, $-\log_{10}$ scale) are plotted against the expected $p$-values ($x$-axis, $-\log_{10}$ scale). The multiple testing correction threshold for significance following Bonferroni correction is denoted by a dotted line. The bottom panel shows the same data as the top panel but excluding the 30 interactions that were significant at the Bonferroni level in the replication datasets. The shaded grey area represents the 5\% confidence interval for the expected distribution of $p$-values. Dark blue points represent $p$-values that exceed the confidence interval, light blue are within the confidence interval.}
	\label{fig:qqMeta}
\end{figure}

\begin{figure}[H]
	\centering
	% \includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes). Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data The remaining 89 interactions that did not replicate are depicted in grey. It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations. An interactive version of this graph can be found here: \url{http://kn3in.github.io/detecting_epi/}}
	\label{fig:fireworks}
\end{figure}

@@

\setcounter{page}{1}
\section*{Supplementary Figures}
\setcounter{figure}{0}
\makeatletter 
\renewcommand{\thefigure}{S\@arabic\c@figure} 
\makeatletter

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{qqMetaNonsig}
	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets, excluding the 30 points significant at the Bonferroni level} The right panel (Null) shows the interaction $p$-values from a meta analysis across two independent datasets on 434 SNP pairs where one SNP has a marginal effect. The left panel (Empirical) shows the interaction $p$-values from the 404 putative interactions that were not significant at the Bonferroni correction threshold. Dark blue points represent $p$-values that surpass the 2.5\% FDR level, as in Figure \ref{fig:qqMeta}.}
\label{fig:qqMetaNonsig}
\end{figure}
\clearpage
